---
title: "LakeColor Model Interpretability"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(tidyverse)
library(xgboost)
library(Metrics)
library(gridExtra)
knitr::opts_chunk$set(echo = F, warning = F, error = F, message = F)

```


```{r}
load('data/modelInput_20210302.RData')

fui.lookup <- tibble(dWL = c(471:583), fui = NA)

fui.lookup$fui[fui.lookup$dWL <= 583] = 21
fui.lookup$fui[fui.lookup$dWL <= 581] = 20
fui.lookup$fui[fui.lookup$dWL <= 579] = 19
fui.lookup$fui[fui.lookup$dWL <= 577] = 18
fui.lookup$fui[fui.lookup$dWL <= 575] = 17
fui.lookup$fui[fui.lookup$dWL <= 573] = 16
fui.lookup$fui[fui.lookup$dWL <= 571] = 15
fui.lookup$fui[fui.lookup$dWL <= 570] = 14
fui.lookup$fui[fui.lookup$dWL <= 569] = 13
fui.lookup$fui[fui.lookup$dWL <= 568] = 12
fui.lookup$fui[fui.lookup$dWL <= 567] = 11
fui.lookup$fui[fui.lookup$dWL <= 564] = 10
fui.lookup$fui[fui.lookup$dWL <= 559] = 9
fui.lookup$fui[fui.lookup$dWL <= 549] = 8
fui.lookup$fui[fui.lookup$dWL <= 530] = 7
fui.lookup$fui[fui.lookup$dWL <= 509] = 6
fui.lookup$fui[fui.lookup$dWL <= 495] = 5
fui.lookup$fui[fui.lookup$dWL <= 489] = 4
fui.lookup$fui[fui.lookup$dWL <= 485] = 3
fui.lookup$fui[fui.lookup$dWL <= 480] = 2
fui.lookup$fui[fui.lookup$dWL <= 475 & fui.lookup$dWL >470] = 1
fui.lookup$dWL <- as.numeric(fui.lookup$dWL)

modelInput <- modelInput %>% 
  mutate(mode = round(mode)) %>%
  left_join(fui.lookup, by = c('mode' = 'dWL'))
## Identify features to use as predictors
#names(modelInput)

#c("mode", "bw","dwStd","interQ", "unimodal", "group"   "class")

target <- 'dwStd'
if(target == 'fui'){
  modelInput <- modelInput %>% filter(!is.na(fui))
}


feats <- c("lakeMeanOcc", "mean_2m_air_temperature_mean","mean_2m_air_temperature_stdDev", 
           "total_precipitation_mean", "total_precipitation_stdDev", "depthMean", "volumn",
           "elevation", "shoreLength", "polygonArea", "winter_temperature_mean", 
           "winter_total_precipitation_mean", "spring_temperature_mean", 
           "spring_total_precipitation_mean", "summer_temperature_mean",
           "summer_total_precipitation_mean", "fall_temperature_mean", 
           "fall_total_precipitation_mean")

#colSums(is.na(modelInput %>% select(feats)))
#Predictors = `r feats`

```


## First we'll train our model.  

Here, I let the model choose which predictors are most important.  This works for xgboost a little better than for random forests because its not as susceptible to misplacing importance due to collinearity.  We'll look at feature importance using a couple different metrics all the same. 

Below, we train our model until the validation rmse stops improving. This avoids overfitting. We check the RMSE using both 5 fold cross validation and using holdout test data to get two different measures of model performance.

#### Model target = `r target` 

Note: Ultimately, we should probably decide on only one modeling approach.  All the interpretability metrics I use here are model agnostic, meaning we can apply them to random forest, xgboost, etc.

```{r}
set.seed(2423)
train <- modelInput %>% sample_frac(.7)
test <- modelInput %>% filter(!id %in% train$id)

dtrain <- xgb.DMatrix(data = as.matrix(train[,feats]), label = train[,target][[1]])
dtest <- xgb.DMatrix(data = as.matrix(test[,feats]), label = test[,target][[1]])


params <- list(booster = "gbtree", objective = "reg:squarederror", eta=0.3, gamma=0, max_depth=3, min_child_weight=1, subsample=1, colsample_bytree=1)

## Do a quick cv to check ideal number of folds
xgbcv <- xgb.cv(params = params, data = dtrain, nrounds = 1000, nfold = 5, showsd = T, stratified = T, print_every_n = 25, early_stopping_round = 10, maximize = F)

xgb.naive <- xgb.train(params = params, data = dtrain, nrounds = 1000, watchlist = list(train = dtrain, val = dtest), print_every_n = 25, early_stopping_rounds = 10, maximize = F)
```

#### We want our CV RMSE to roughly equal our holdout RMSE, if holdout is much lower we might be overfitting.

#### `r paste0('Naive Model Hold-out RMSE = ', xgb.naive$best_score)`
#### `r paste0('CV RMSE = ', xgbcv$evaluation_log$test_rmse_mean[xgbcv$best_iteration])`
  
Note: We still need to think about is how we want to portray our validation. The non-linear aspects of dominant wavelength makes it a little tricky.

```{r, eval = F}
    
preds <- test %>% mutate(predicted = predict(xgb.naive, dtest))
                         
plotdf <- preds %>% select(all_of(target), 'predicted') %>%
  mutate_all(round)

plotdf$residual <- unlist(plotdf[1]-plotdf[2])

ggplot(plotdf, aes(x = residual)) + geom_density()

ggplot(preds, aes_string(x = target, y = 'predicted')) +
  geom_hex() +
  scale_fill_viridis_c(trans = 'log10') +
  geom_abline(color = 'red')

ggplot(plotdf, aes_string(x = target, y = 'predicted')) +
  geom_bin2d() +
  scale_fill_viridis_c(trans = 'log10') +
  geom_abline(color = 'red') +
  labs(x = 'actual', y = 'predicted', title = target)
```

## Now lets look at some interpretability metrics, first feature importance and Accumulated Local Effects (ALE) plots

Here, feature importance is permutation feature importance, or the increase in error when the values of a predictor are randomly shuffled (i.e. when we add significant noise to a given predictor).  

```{r}
library(iml)

pred <- function(model, newdata){
  predict(model, xgb.DMatrix(as.matrix(newdata)))
}

predictor <- Predictor$new(xgb.naive, data = train[,feats], y = train[,target][[1]], predict.function = pred)

featureImp <- FeatureImp$new(predictor, loss = 'mse')
plot(featureImp)
effs <- FeatureEffects$new(predictor)
check <- do.call('rbind', effs$results)
```

ALE plots describe the average influence of a predictor on the final prediction along a localized window of values.  These plots are simple to interpret, fast to calculate, and aren't impacted collinearity in the predictor space.  If the ALE value is above zero, it means the feature has a positive impace on model predictions at the given value along the x-axis. If it's below 0, it has a negative impact. The distributions above each plot represent the distibution of values (5-95th percentile) we have in our training data (not sure why the plot alignment is a little off, this can be fixed down the road). 

For a summary of ALE plots, see https://christophm.github.io/interpretable-ml-book/ale.html.

```{r}
alePlotter <- function(feature){
  perc5 = quantile(train[[feature]],.04)
  perc95 = quantile(train[[feature]],0.96)
  
  p1 <- ggplot(train, aes_string(x = feature)) + geom_density(adjust = 4, fill = 'grey70') +
    xlim(perc5,perc95) + 
    theme_classic() +
    theme(axis.text = element_text(color = 'transparent'),
          axis.title = element_blank(),
          #axis.title.y = element_text(color = 'transparent'),
          axis.ticks = element_blank(),
          axis.line = element_line(color = 'transparent'),
          plot.margin = margin(0,-1,-1,-1))
  
  #if(feature == 'dWL'){p1 = p1 + labs(tag = 'b)')}
  
  p2 <- check %>% filter(.feature == feature, .borders >= perc5, .borders <= perc95) %>%
    ggplot(.,aes(x= .borders, y = .value)) +
    geom_line() +
    geom_point() +
    xlim(perc5,perc95) +
    geom_hline(aes(yintercept = 0), color = 'red')+
    facet_wrap(~.feature) +
    theme_bw() +
    theme(axis.title = element_blank(),
          plot.margin = margin(-3,0,0,0))
  #if(feature == 'dWL'){p2 = p2 + labs(tag = 'b)')}
  arrangeGrob(p1,p2, nrow = 2, heights = c(.3,1))
  }


p1 <- alePlotter(featureImp$results$feature[1])  
p2 <- alePlotter(featureImp$results$feature[2])
p3 <- alePlotter(featureImp$results$feature[3])
p4 <- alePlotter(featureImp$results$feature[4])
p5 <- alePlotter(featureImp$results$feature[5])
p6 <- alePlotter(featureImp$results$feature[6])
p7 <- alePlotter(featureImp$results$feature[7])
p8 <- alePlotter(featureImp$results$feature[8])
p9 <- alePlotter(featureImp$results$feature[9])
p10 <- alePlotter(featureImp$results$feature[10])
p11 <- alePlotter(featureImp$results$feature[11])
p12 <- alePlotter(featureImp$results$feature[12])


grid.arrange(p1,p2,p3,p4,p5,p6,p7,p8,p9,p10,p11,p12, nrow = 4, left = 'Accumulated Local Effect (nm)', bottom = 'Feature Value')
```

### Surrogate Trees can help us identify what variable splits and thresholds might be most important.
These are similar to the results that Xiao has presented, except that they're based on the *predictions* of our machine learning model and not the actual *observed* values in our dataset.  This means that it's telling us important splits/thresholds in the model itself. Here I limit them to a depth of 2, but this is adjustable. 

```{r}
tree <- TreeSurrogate$new(predictor, maxdepth = 2)
plot(tree$tree)
```

### Finally, we'll look at SHAP (Shapely Additive exPlanations).
These are similar to ALE plots, but show the distribution of feature effects across all observations instead of just averaged across a small window.  Additionally, they can tell us the feature contributions to any *individual* prediction.  The methods are a little more complicated, but a simplified explanation is that SHAP builds iterative local models at each of our observations. For more info, see https://christophm.github.io/interpretable-ml-book/shap.html

First: we'll look at overall SHAP results for our nine most important features. For SHAP, importance is calculated as those predictors that have the highest cumulative impact across the local models.

```{r}
xgb.plot.shap(data = as.matrix(train[,feats]), top_n = 9, n_col = 3, model = xgb.naive)
```

Then, we'll look at the SHAP summary plot for all predictors. Here, features are ordered by their SHAP importance. Each point on the summary plot is a Shapley value for a feature in one of our observations. The x-axis is it's Shapley value (it's contribution to the model prediction). The color represents the relative value of the feature from low to high.

```{r}
library(SHAPforxgboost)
shap_long <- shap.prep(xgb_model = xgb.naive, X_train = as.matrix(train[,feats]))
shap.plot.summary(shap_long)
```

With SHAP, we can take any individual observation and see how each input feature contributes to the final prediction as shown below for the first observation in our training dataset. Phi, along the x axis, denotes the impact of each feature to the final prediction. This allows us to look at any individual lake and see how our model is predicting its color.

```{r}
shap <- Shapley$new(predictor, x.interest = train[,feats][1,])
plot(shap)
```
